Learning Causally-Generated Stationary Time Series

### 2017-02-23 doc/paper.tex
Various updates

A widely-used model for a stationary time series $f:\R \to \R$ is the Gaussian process, which constructs a prior over functions by assuming that any finite collection of function values $f(t_1),\ldots,f(t_n)$ is multivariate-Gaussian distributed. An important modelling decision in using a Gaussian process is the choice of covariance   $k_f(t,t')$ between any two function values $f(t)$ and $f(t')$---$k_f$ is also called the \textit{kernel} of $f$. This choice of kernel entirely\footnote{The mean of any function value is without loss of generality assumed to be zero. \cite{Rasmussen:2006:Gaussian_Processes}} determines the nature of $f$. Hence much work has been done in developing flexible kernels \cite{Duvenaud:2014:Automatic_Construction,Wilson:2013:Spectral_Mixture,Tobar:2015:Learning_Stationary,Tobar:2015:Inter-Domain_Inducing}; most notably, \citet{Tobar:2015:Learning_Stationary} present the Gaussian Process Convolution Model (GPCM), which parametrises the kernel with \textit{another} Gaussian process.

@@

\section{Modelling Causally-Generated Stationary Time Series}
Consider modelling the stationary time series $f$. Since many dynamical systems in nature can accurately be described by an initial value problem or a linear system, we let $f$ be the solution of a time-invariant linear initial value problem with causal Green's function $h$ and forcing function $x$, or equivalently the system response of a time-invariant linear system with causal impulse response $h$ and excitation $x$. It then holds that
\begin{align} \label{eq:model}
    f(t) = \int^t h(t- \tau)x(\tau)\id{\tau}.
\end{align}
We denote integration from negative infinity or to positive infinity by respectively omitting the lower or upper limit of the integral.
Note that $f(t)$ depends on $x(\tau)$ only for $\tau \le t$; this reveals \cref{eq:model}'s causal nature.

@@

Since $h$ consitutes an unknown function, we choose to model it using a Gaussian process; that is, we let $h \sim \mathcal{GP}(0,k_h(t,t'))$. In choosing $k_h(t,t')$, we require that $f$ has finite power, since every real-world signal has so as well. This can be achieved by letting $h$ decay to zero at infinity sufficiently quickly \cite{Tobar:2015:Learning_Stationary}: let $h\cond g = w g$ where $w(t)= \exp(- \alpha t^2)$ and $g \sim \mathcal{GP}(0,k_g(t-t'))$ has finite power; in that case,
\begin{align*}
    \mathbb{V}[f(t)]
    &= \int_{0}\exp(-2 \alpha t^2)\mathbb{V}[g(\tau)]\id{\tau} \\
    &\le k_g(0) \int_0 \exp(- 2 \alpha t^2) \id{\tau} < \infty,
\end{align*}
which could be infinite otherwise.
Equivalently, we let $k_h(t,t')=\exp(- \alpha (t^2 + t^{\prime 2}))k_g(t-t')$. To retain flexibility in the prior on $h$, we let $k_g$ be an exponentiated square. We thus have that $k_h(t,t')=\exp(- \alpha (t^2 + t^{\prime 2}) - \gamma(t-t')^2)$. Note that $\alpha$ determines $k_f$'s length scale and $\gamma$ its flexibility.

@@

The CPGCM is similar to the latent force model presented by \citet{Alvarez:2009:Latent_Force_Models}: whereas we let $h$ be free-form and let $x$ be white noise, \citet{Alvarez:2009:Latent_Force_Models} specify $h$ deterministically and let $x$ be free-form.

\subsection{Sampling from the CGPCM}
Sampling from the CGPCM is difficult because the integrals in \cref{mod:cgpcm,mod:cgpcm2} depend on the entirety of $h$ and $x$. To resolve this issue, we follow \citet{Tobar:2015:Learning_Stationary} and approximate \cref{eq:kernel} using quadrature \cite{Minka:2000:Quadrature_GP} by conditioning on finitely many values $u=(h(t_{u,1}),\ldots,h(t_{u,n_u}))\sim \mathcal{N}(0,K_u)$ of $h$:
\begin{align*}
    k_{f\cond h}(r)
    &\approx \mathbb{E}[k_{f\cond h}(r)\cond u]
    = \int_{0} \mathbb{E}[h(|r| + \tau) h(\tau)\cond u] \id{\tau} \\
    &= \int_0 k_h(|r| + \tau, \tau) \id{\tau} \\
    &\phan{=}+ \tr M^u \int_{0} k_h(t_u, |r|+ \tau)k_h(\tau, t_u^T) \id{\tau}
\end{align*}
where $M^{u}=K_u^{-1}uu^T K_u^{-1}-K_u$. Unlike the GPCM, the kernel approximation $\mathbb{E}[k_{f\cond h}(r)\cond u]$ of the CGPCM will not be a mixture of Gaussians, but a more complicated expression instead.

@@

\section{Inference}
Let $y(t)\cond \sim \mathcal{N}(f(t),\sigma^2)$ for all $t$ be a noisy version of $f$. Then, given some observations $e=(y(t_1),\ldots,y(t_n))$ of $y$, we wish to compute $p(f\cond e)$.

We learn $h$ and $x$ through inducing points $u=(h(t_{u,1}),\ldots,h(t_{u,n_u}))\sim \mathcal{N}(0,K_u)$ and $z=(s(t_{z,1}),\ldots,s(t_{z,n_z}))\sim \mathcal{N}(0,K_z)$ for respectively the processes $h$ and $s=T[x]$ where $T$ is some linear interdomain transformation \cite{Titsias:2009:Variational_Learning,Lazaro-Gredilla:2009:Inter-Domain_Gaussian_Processes_for_Sparse,Alvarez:2010:Efficient_Multioutput_Gaussian_Processes_Through,Tobar:2015:Learning_Stationary}. Concretely, let the \textit{mean-field} (MF) approximation $q(f,u,z)=p(f\cond u, z)q(u)q(z)$ of $p(f,u,z\cond e)$ be such that it is closest to $p(f,u,z\cond e)$ in Kullback-Leibler divergence.
% \begin{align*}
%     (q(u),q(z))&= \underset{(q(u),q(z))}{\argmin}\KL(q(f,u,z)\|p(f,u,z\cond e)).
% \end{align*}

Rearranging
\begin{align*}
    &\log p(e) - \KL(q(f,u,z)\|p(f,u,z\cond e)) \\
    &\quad= \ll\la \log \frac{p(e\cond f)\cancel{p(f\cond u, z)}p(u)p(z)}{\cancel{p(f\cond u, z)}q(u)q(z)} \rr\ra_{q(f,u,z)} \\
    &\quad= \underbrace{\la \log p(e\cond f) \ra_{q(f,u,z)}}_{\text{reconstruction cost}} \\
    &\quad\phan{=}- \underbrace{\ll(\ll\la\log\frac{q(u)}{p(u)}\rr\ra_{q(u)} + \ll\la\log\frac{q(z)}{p(z)}\rr\ra_{q(z)}\rr)}_{\text{divergence from prior}} \\
    &\quad=\mathcal{L}[q(u),q(z)]
\end{align*}
shows that can find $q(u)$ and $q(z)$ by maximising $\mathcal{L}$. Since $\log p(e)\ge\mathcal{L}$, $\mathcal{L}$ is called the \textit{evidence lower bound} (ELBO). Observe that maximisation of $\mathcal{L}$ attempts to explain the data well whilst not diverging too far from the model prior.

@@

We can, however, significantly speed up the optimisation by solving for either $q(u)$ or $q(z)$ analytically. Substituting the optimal form $q(z)$ back in $\mathcal{L}$ yields that
\begin{align}
    &\mathcal{L}^*[q(u)] \nonumber\\
    &\quad=\max_{q(z)} \mathcal{L}[q(u),q(z)]\nonumber\\
    &\quad= \log \int p(z)\exp \la \log p(e\cond f) \ra_{p(f\cond u, z)q(u)} \id{z} \nonumber\\
    &\quad\phan{=}-\ll\la\log\frac{q(u)}{p(u)}\rr\ra_{q(u)},\label{eq:saturated_elbo}
\end{align}
which is computed in \cref{app:saturated_elbo}. We optimise this saturated lower bound $\mathcal{L}^*$ to yield $q(u)$ and then obtain $q(z)$ through \cref{eq:qz}.

@@

Following an argument similar to the one previously presented, we derive the corresponding ELBO:
\begin{align*}
    &\mathcal{L}[q(u),q(z\cond u)] \\
    &\quad= \la \log p(e\cond f) \ra_{q(f,u,z)}- \ll\la\log\frac{q(u)q(z\cond u)}{p(u)p(z)}\rr\ra_{q(u)q(z\cond u)}.
\end{align*}
Again, setting the variations $\delta \mathcal{L} / \delta q(u)$ and $\delta \mathcal{L} / \delta q(z\cond u)$ to zero and solving for for respectively $q(u)$ and $q(z\cond u)$ yields that
\begin{align}
    q(u) &\propto p(u) \int p(z) \exp\la\log p(e\cond f)\ra_{p(f\cond u, z)}\id{z}, \label{eq:qu-smf} \\
    q(z\cond u) &\propto p(z)\exp\la \log p(e\cond f)\ra_{p(f\cond u, z)}, \label{eq:qz-smf}
\end{align}
which are computed in \cref{app:computation_quz}.
As opposed to \cref{eq:qu,eq:qz}, \cref{eq:qu-smf,eq:qz-smf} are uncoupled, but $q(u)$'s moments are now intractable. We can, however, evaluate $q(u)$. We therefore use elliptical slice sampling (ESS) \cite{Murray:2010:Elliptical_Slice_Sampling} to sample from $q(u)$ and use these samples to approximate $q(f, u, z)$. To help mixing the Markov chain, we initialise the sampler with the MF approximation.

\subsection{Interdomain Transformation}
The interdomain transformation $T$ remains yet unspecified. We let
\begin{align*}
    T[x](t)=\int r(t- \tau)x(\tau) \id{\tau}
\end{align*}
where $r$ is some filter. In choosing $r$, we should make sure that $s=T[x]$ has power at the majority of frequencies present in the signal that we aim to model.

For baseband signals, we follow \citet{Tobar:2015:Learning_Stationary} and let $r(t)=\exp(-\omega t^2)$. For signals that have power at all frequencies, such as nondifferentiable signals, we let $r(t)=\exp(- \omega t^2)H(t)$ where $H$ denotes the Heaviside step function; namely, the step at $t=0$ spreads power throughout the whole spectrum. Note that this choice of $T$ yields a \textit{causal interdomain transformation}.

\section{Experiments}
We used TensorFlow \cite{Abadi:2016:TensorFlow_A_System_for_Large-Scale} to implement the CGPCM.\footnote{The implementation can be found at \url{https://github.com/wesselb/cgpcm}.} The main issue in implementing the CGPCM is that computation of the matrices in \cref{app:moments_f} requires extensive evaluation of the bivariate normal cumulative density function (BNCDF) \cite{Genz:2004:Numerical_Computation_of_Rectangular_Bivariate}. Since the BNCDF is not available in TensorFlow, we used the implementation of the BNCDF from the R package \texttt{pbivnorm}\footnote{See \url{https://github.com/brentonk/pbivnorm}.} to implement the BNCDF and its gradient in TensorFlow. We furthermore implementated an algorithm that symbolically solves the integrals in \cref{app:moments_f} and constructs the corresponding matrices afterwards; namely, manually solving these expressions  otherwise would be tedious and error prone.

\subsection{Ornstein-Uhlenbeck Process}
The Ornstein-Uhlenbeck process is a noisy relaxation process that can be formulated as a Gaussian process with kernel of the form $k(r)=\exp(- \alpha |r|)$ \cite{Rasmussen:2006:Gaussian_Processes}. This corresponds to the particular case of the CGPCM where $h(t)=\sqrt{2 \alpha} \exp(- \alpha |t|)$.

@@

@@

\section{Discussion}
We have presented the CGPCM as a model for causally-generated stationary time series. Inference is performed by finding an initial approximation through a variational mean-field approach, and then refining this approximation through a variational structured mean-field approach combined with elliptical slice sampling. The GPCM has been validated {\color{red} discussion of results of experiments here.}

In future research on the CGPCM, various directions can be explored: First, the model could be extented to multidimensional input and output spaces by following a construction similar to that by \citet{Bruinsma:2016:GGPCM}. Second, the application of more efficient sparse approximation for $h$ and $x$, such as the tree-structured approximation \cite{Bui:2014:Tree-Structured_Gaussian}, would scale the CGPCM to larger data sets. Third, to deal with signals that are band-limited, but not necessarily baseband, the use of an harmonic interdomain transformation \cite{Tobar:2015:Inter-Domain_Inducing} and an harmonic extension of the causal interdomain transformation could be explored. Fourth, in the low noise case, ESS has a hard time exploring $q(u)$; therefore, in that case, afternative sampling algorithms such as Hamiltonian Monte-Carlo \cite{Neal:2012:MCMC_Using_Hamiltonian_Dynamics} could be used. Finally, using an approach similar to that by \citet{Lazaro-Gredilla:2013:Variational_Inference_for_Mahalanobis_Distance}, Bayesian inference in the hyperparameters could be attempted.

@@

\begin{align*}
    &\la f(t) f(t') \ra_{p(f\cond u,z)} \\
    &\quad= \int^t\!\!\!\!\int^{t'} \la h(t- \tau) h(t' - \tau')\ra_{p(h\cond u)} \\
    &\quad\phan{=\int^t\!\!\!\!\int^{t'}} \la x(\tau) x(\tau') \ra_{p(x\cond z)}\id{\tau'}\id{\tau} \\
    &\quad= \int^t\!\!\!\!\int^{t'} ( k_h(t- \tau,t' - \tau') \\
    &\quad\phan{=\int^t\!\!\!\!\int^{t'} (}+ k_h(t- \tau, t_u^T) M^u k_h(t_u, t' - \tau')) \\
    &\quad\phan{=\int^t\!\!\!\!\int^{t'}}( k_x(\tau,\tau') + k_{xs}(\tau, t_z^T) M^z k_{sx}(t_z, \tau')) \id{\tau'}\id{\tau} \\
    &\quad= a(t,t') + \tr M^u A^{h}(t,t') + \tr M^z A^{x}(t,t') \\
    &\quad\phan{=} + \tr M^u A^{hx}(t) M^z A^{xh}(t')
\end{align*}
where $M^u=K_u^{-1}uu^T K_u^{-1}-K_u^{-1}$, $M^z=K_z^{-1}zz^TK_z^{-1}-K_z^{-1}$, and
\begin{align*}
    a(t,t')&=\int^t\!\!\!\!\int^{t'} k_h(t- \tau,t' - \tau') k_x(\tau,\tau') \id{\tau'}\id{\tau} \\
    &=\int^{t \land t'} k_h(t- \tau,t' - \tau)\id{\tau}, \\
    A^{h}(t,t')&=\int^t\!\!\!\!\int^{t'} k_h(t_{u}, t' - \tau') k_x(\tau,\tau') \\
    &\phan{=\int^t\!\!\!\!\int^{t'}}\, k_h(t- \tau, t_{u}^T) \id{\tau'}\id{\tau}\\
    &=\int^{t \land t'} k_h(t_{u}, t' - \tau)k_h(t- \tau, t_{u}^T) \id{\tau}  \\
    A^{x}(t,t')&=\int^t\!\!\!\!\int^{t'} k_{sx}(t_z, \tau') k_h(t- \tau,t'-\tau') \\
    &\phan{=\int^t\!\!\!\!\int^{t'}}\, k_{xs}(\tau, t_z^T) \id{\tau'}\id{\tau}.
\end{align*}
Rearranging, we arrive at $\la f(t) \ra = u^T  A^{hx}(t) z$ and
\begin{align*}
    &\la f(t) f(t') \ra - \la f(t) \ra \la f(t') \ra \\
    &\quad = b(t,t') + u^T B^{h}(t,t') u + z^T B^{x}(t,t') z
\end{align*}
where the expectation is over $p(f\cond K_u^{-1} u, K_z^{-1} z)$ and
\begin{align*}
    b(t,t) &= a(t,t') - \tr K_u^{-1} A^{h}(t,t') - \tr K_z^{-1}A^{x}(t,t') \\
    &\phan{=} + \tr K_u^{-1}A^{hx}(t)K_z^{-1}A^{xh}(t'), \\
    B^{h}(t,t') &= A^{h}(t,t') - A^{hx}(t)K_z^{-1} A^{xh}(t'), \\
    B^{x}(t,t') &= A^{x}(t,t') - A^{xh}(t)K_u^{-1} A^{hx}(t').
\end{align*}
Finally denote $a(t)=a(t,t)$ and do so for $A^h$, $A^x$, $A^{hx}$, $b$, $B^h$, and $B^x$ as well.

\subsection{Causal Interdomain Transformation}
In the case of the causal interdomain transformation, $A^x$ and $A^{hx}$ reduce to
\begin{align*}
   A^{hx}(t) &=\int^{t\land t_z^T} k_h(t_u,t-\tau) k_{xs}(\tau, t_z^T) \id{ \tau}, \\
   A^{x}(t,t')&=\int^{t \land t_z}\!\!\!\!\int^{t'\land t_z^T} k_{sx}(t_z, \tau') k_h(t- \tau,t'-\tau') \\
    &\phan{=\int^{t \land t_z}\!\!\!\!\int^{t'\land t_z^T}}\, k_{xs}(\tau, t_z^T) \id{\tau'}\id{\tau}.
\end{align*}

>>>>

A widely-used model for a stationary time series $f:\R \to \R$ is the Gaussian process, which constructs a prior over functions by assuming that any finite collection of function values $f(t_1),\ldots,f(t_n)$ is multivariate-Gaussian distributed. An important modelling decision in using a Gaussian process is the choice of covariance   $k_f(t,t')$ between any two function values $f(t)$ and $f(t')$---$k_f$ is also called the \textit{kernel} of $f$. This choice of kernel entirely\footnote{The mean of the function is assumed to be zero without loss of generality. \cite{Rasmussen:2006:Gaussian_Processes}} determines the nature of $f$. Hence much work has been done in developing flexible kernels \cite{Duvenaud:2014:Automatic_Construction,Wilson:2013:Spectral_Mixture,Tobar:2015:Learning_Stationary,Tobar:2015:Inter-Domain_Inducing}; most notably, \citet{Tobar:2015:Learning_Stationary} present the Gaussian Process Convolution Model (GPCM), which parametrises the kernel with \textit{another} Gaussian process.

@@

\section{Modelling Causally-Generated Stationary Time Series}
Consider modelling the stationary time series $f$. Since many dynamical systems in nature can accurately be described by an initial value problem or a linear system, we let $f$ be the solution of a time-invariant linear initial value problem with causal Green's function $h$ and forcing function $x$, or equivalently the system response of a time-invariant linear system with causal impulse response $h$ and excitation $x$. It then holds that
\begin{align} \label{eq:model}
    f(t) = \int^t h(t- \tau)x(\tau)\id{\tau}.
\end{align}
We denote integration from negative infinity or to positive infinity by omitting the respective limit from the integral.
Note that $f(t)$ depends on $x(\tau)$ only for $\tau \le t$; this reveals \cref{eq:model}'s causal nature.

@@

Since $h$ consitutes an unknown function, we choose to model it using a Gaussian process; that is, we let $h \sim \mathcal{GP}(0,k_h(t,t'))$. In choosing $k_h(t,t')$, we require that $f$ has finite power, since every real-world signal has so as well. Following \citet{Tobar:2015:Learning_Stationary}, this can be achieved by letting $h$ decay to zero at infinity sufficiently quickly: let $h\cond g = w g$ where $w(t)= \exp(- \alpha t^2)$ and $g \sim \mathcal{GP}(0,k_g(t-t'))$ has finite power; in that case,
\begin{align*}
    \mathbb{V}[f(t)]
    &= \int_{0}\exp(-2 \alpha t^2)\mathbb{V}[g(\tau)]\id{\tau} \\
    &\le k_g(0) \int_0 \exp(- 2 \alpha t^2) \id{\tau} < \infty,
\end{align*}
which could be infinite otherwise.
Equivalently, we let $k_h(t,t')=\exp(- \alpha (t^2 + t^{\prime 2}))k_g(t-t')$. To retain flexibility in the prior on $h$, we let $k_g$ be an exponentiated square. We thus have that $k_h(t,t')=\exp(- \alpha (t^2 + t^{\prime 2}) - \gamma(t-t')^2)$. Note that $\alpha$ determines $k_f$'s length scale and $\gamma$ its flexibility.

@@

The CPGCM is similar to the latent force model presented by \citet{Alvarez:2009:Latent_Force_Models}; however, whereas we let $h$ be free-form and let $x$ be white noise, \citet{Alvarez:2009:Latent_Force_Models} specify $h$ deterministically and let $x$ be free-form.

\subsection{Sampling from the CGPCM}
Sampling from the CGPCM is difficult because the integrals in \cref{mod:cgpcm,mod:cgpcm2} depend on the entirety of $h$ and $x$. To resolve this issue, we follow \citet{Tobar:2015:Learning_Stationary} and approximate \cref{eq:kernel} using quadrature \cite{Minka:2000:Quadrature_GP} by conditioning on finitely many values $u=(h(t_{u,1}),\ldots,h(t_{u,n_u}))\sim \mathcal{N}(0,K_u)$ of $h$:
\begin{align*}
    k_{f\cond h}(r)
    &\approx \mathbb{E}[k_{f\cond h}(r)\cond u]
    = \int_{0} \mathbb{E}[h(|r| + \tau) h(\tau)\cond u] \id{\tau} \\
    &= \int_0 k_h(|r| + \tau, \tau) \id{\tau} \\
    &\phan{=}+ \tr M^u \int_{0} k_h(t_u, |r|+ \tau)k_h(\tau, t_u^T) \id{\tau}
\end{align*}
where $M^{u}=K_u^{-1}uu^T K_u^{-1}-K_u$. Unlike the GPCM, the kernel approximation $\mathbb{E}[k_{f\cond h}(r)\cond u]$ of the CGPCM will not simply be a mixture of Gaussians; it will be a more complicated expression instead.

@@

\section{Inference}
Let $y(t)\cond \sim \mathcal{N}(f(t),\sigma^2)$ for all $t$ be a noisy version of $f$. Given some observations $e=(y(t_1),\ldots,y(t_n))$ of $y$, we wish to compute $p(f,h,x\cond e)$.

We learn $h$ and $x$ through inducing points \cite{Titsias:2009:Variational_Learning} $u=(h(t_{u,1}),\ldots,h(t_{u,n_u}))\sim \mathcal{N}(0,K_u)$ and $z=(s(t_{z,1}),\ldots,s(t_{z,n_z}))\sim \mathcal{N}(0,K_z)$ for respectively the processes $h$ and $s=T[x]$ where
\begin{align*}
    T[x](t)=\int r(t- \tau)x(\tau) \id{\tau}
\end{align*}
is some interdomain transformation with filter $r$ \cite{Lazaro-Gredilla:2009:Inter-Domain_Gaussian_Processes_for_Sparse,Alvarez:2010:Efficient_Multioutput_Gaussian_Processes_Through,Tobar:2015:Learning_Stationary}. In choosing $r$, we make sure that $s=T[x]$ has power at the majority of frequencies present in the signal that we aim to model. We follow \citet{Tobar:2015:Learning_Stationary} and let $r(t)=\exp(-\omega t^2)$, which fits baseband signals.

Let the \textit{mean-field} (MF) approximation $q(f,u,z)=p(f\cond u, z)q(u)q(z)$ of $p(f,u,z\cond e)$ be such that it is closest to $p(f,u,z\cond e)$ in Kullback-Leibler divergence. Rearranging
\begin{align*}
    &\log p(e) - \KL(q(f,u,z)\|p(f,u,z\cond e)) \\
    &\quad= \ll\la \log \frac{p(e\cond f)\cancel{p(f\cond u, z)}p(u)p(z)}{\cancel{p(f\cond u, z)}q(u)q(z)} \rr\ra_{q(f,u,z)} \\
    &\quad= \underbrace{\la \log p(e\cond f) \ra_{q(f,u,z)}}_{\text{reconstruction cost}} \\
    &\quad\phan{=}- \underbrace{\ll(\ll\la\log\frac{q(u)}{p(u)}\rr\ra_{q(u)} + \ll\la\log\frac{q(z)}{p(z)}\rr\ra_{q(z)}\rr)}_{\text{divergence from prior}} \\
    &\quad=\mathcal{L}[q(u),q(z)]
\end{align*}
shows that we can find $q(u)$ and $q(z)$ through maximising $\mathcal{L}$. Since $\log p(e)\ge\mathcal{L}$, $\mathcal{L}$ is called the \textit{evidence lower bound} (ELBO). Observe that maximisation of $\mathcal{L}$ attempts to explain the data well whilst not diverging too far from the model prior.

@@

We can, however, significantly speed up the optimisation by solving for either $q(u)$ or $q(z)$ analytically. Substituting the optimal form of $q(z)$ back in $\mathcal{L}$ yields that
\begin{align}
    &\mathcal{L}^*[q(u)] \nonumber\\
    &\quad=\max_{q(z)} \mathcal{L}[q(u),q(z)]\nonumber\\
    &\quad= \log \int p(z)\exp \la \log p(e\cond f) \ra_{p(f\cond u, z)q(u)} \id{z} \nonumber\\
    &\quad\phan{=}-\ll\la\log\frac{q(u)}{p(u)}\rr\ra_{q(u)},\label{eq:saturated_elbo}
\end{align}
which is computed in \cref{app:saturated_elbo}. We optimise this saturated lower bound $\mathcal{L}^*$ to yield $q(u)$ and then obtain $q(z)$ through \cref{eq:qz}.

@@

Following an argument similar, we again derive the corresponding ELBO:
\begin{align*}
    &\mathcal{L}[q(u),q(z\cond u)] \\
    &\quad= \la \log p(e\cond f) \ra_{q(f,u,z)}- \ll\la\log\frac{q(u)q(z\cond u)}{p(u)p(z)}\rr\ra_{q(u)q(z\cond u)}.
\end{align*}
Again, setting the variations $\delta \mathcal{L} / \delta q(u)$ and $\delta \mathcal{L} / \delta q(z\cond u)$ to zero and solving for for respectively $q(u)$ and $q(z\cond u)$ yields that
\begin{align}
    q(u) &\propto p(u) \int p(z) \exp\la\log p(e\cond f)\ra_{p(f\cond u, z)}\id{z}, \label{eq:qu-smf} \\
    q(z\cond u) &\propto p(z)\exp\la \log p(e\cond f)\ra_{p(f\cond u, z)}, \label{eq:qz-smf}
\end{align}
which are computed in \cref{app:computation_quz}.
As opposed to \cref{eq:qu,eq:qz}, \cref{eq:qu-smf,eq:qz-smf} are uncoupled, but $q(u)$'s moments are now intractable. We can, however, evaluate $q(u)$. We therefore use elliptical slice sampling (ESS) \cite{Murray:2010:Elliptical_Slice_Sampling} to sample from $q(u)$ and use these samples to approximate $q(f, u, z)$. To help mixing the Markov chain, we initialise the sampler with the MF approximation.

\section{Experiments}
We used TensorFlow \cite{Abadi:2016:TensorFlow_A_System_for_Large-Scale} to implement the CGPCM.\footnote{The implementation can be found at \url{https://github.com/wesselb/cgpcm}.} The main issue in implementing the CGPCM is that computation of the matrices in \cref{app:moments_f} requires the bivariate normal cumulative density function (BNCDF) \cite{Genz:2004:Numerical_Computation_of_Rectangular_Bivariate}. Since the BNCDF is not available in TensorFlow, we used the implementation of the BNCDF from the R package \texttt{pbivnorm}\footnote{See \url{https://github.com/brentonk/pbivnorm}.} to implement the BNCDF and its gradient in TensorFlow. We furthermore implementated an algorithm that symbolically solves the integrals in \cref{app:moments_f} and constructs the corresponding matrices afterwards; namely, manually solving these expressions  otherwise would be tedious and error prone.

\subsection{Learning Known Kernels}

@@

% \subsection{Ornstein-Uhlenbeck Process}
% The Ornstein-Uhlenbeck process is a noisy relaxation process that can be formulated as a Gaussian process with kernel of the form $k(r)=\exp(- \alpha |r|)$ \cite{Rasmussen:2006:Gaussian_Processes}. This corresponds to the particular case of the CGPCM where $h(t)=\sqrt{2 \alpha} \exp(- \alpha |t|)$.

@@

\section{Discussion}
We have presented the CGPCM as a model for causally-generated stationary time series. Inference is performed by finding an initial approximation through a variational MF approach, and then refining this approximation through a variational SMF approach combined with ESS. The GPCM has been validated {\color{red} discussion of results of experiments here.}

In future research on the CGPCM, various directions can be explored: First, the model could be extented to multidimensional input and output spaces by following a construction similar to that by \citet{Bruinsma:2016:GGPCM}. Second, the application of more efficient sparse approximation for $h$ and $x$, such as the tree-structured approximation \cite{Bui:2014:Tree-Structured_Gaussian}, would scale the CGPCM to larger data sets. Third, to deal with signals that are band-limited, but not necessarily baseband, the use of an harmonic interdomain transformation \cite{Tobar:2015:Inter-Domain_Inducing} could be explored. Fourth, in the low noise case, ESS has a hard time exploring $q(u)$; therefore, in that case, afternative sampling algorithms such as Hamiltonian Monte-Carlo \cite{Neal:2012:MCMC_Using_Hamiltonian_Dynamics} could be used. Finally, using an approach similar to that by \citet{Lazaro-Gredilla:2013:Variational_Inference_for_Mahalanobis_Distance}, Bayesian inference in the hyperparameters could be attempted.

@@

\begin{align*}
    &\la f(t) f(t') \ra_{p(f\cond u,z)} \\
    &\quad= \int^t\!\!\!\!\int^{t'} \la h(t- \tau) h(t' - \tau')\ra_{p(h\cond u)} \\
    &\quad\phan{=\int^t\!\!\!\!\int^{t'}} \la x(\tau) x(\tau') \ra_{p(x\cond z)}\id{\tau'}\id{\tau} \\
    &\quad= \int^t\!\!\!\!\int^{t'} ( k_h(t- \tau,t' - \tau') \\
    &\quad\phan{=\int^t\!\!\!\!\int^{t'} (}+ k_h(t- \tau, t_u^T) M^u k_h(t_u, t' - \tau')) \\
    &\quad\phan{=\int^t\!\!\!\!\int^{t'}}( k_x(\tau,\tau') + k_{xs}(\tau, t_z^T) M^z k_{sx}(t_z, \tau')) \id{\tau'}\id{\tau} \\
    &\quad= a(t,t') + \tr M^u A^{h}(t,t') + \tr M^z A^{x}(t,t') \\
    &\quad\phan{=} + \tr M^u A^{hx}(t) M^z A^{xh}(t')
\end{align*}
where $M^u=K_u^{-1}uu^T K_u^{-1}-K_u^{-1}$, $M^z=K_z^{-1}zz^TK_z^{-1}-K_z^{-1}$, and
\begin{align*}
    a(t,t')&=\int^t\!\!\!\!\int^{t'} k_h(t- \tau,t' - \tau') k_x(\tau,\tau') \id{\tau'}\id{\tau} \\
    &=\int^{t \land t'} k_h(t- \tau,t' - \tau)\id{\tau}, \\
    A^{h}(t,t')&=\int^t\!\!\!\!\int^{t'} k_h(t_{u}, t' - \tau') k_x(\tau,\tau') \\
    &\phan{=\int^t\!\!\!\!\int^{t'}}\, k_h(t- \tau, t_{u}^T) \id{\tau'}\id{\tau}\\
    &=\int^{t \land t'} k_h(t_{u}, t' - \tau)k_h(t- \tau, t_{u}^T) \id{\tau}  \\
    A^{x}(t,t')&=\int^t\!\!\!\!\int^{t'} k_{sx}(t_z, \tau') k_h(t- \tau,t'-\tau') \\
    &\phan{=\int^t\!\!\!\!\int^{t'}}\, k_{xs}(\tau, t_z^T) \id{\tau'}\id{\tau}.
\end{align*}
Rearranging, we arrive at $\la f(t) \ra = u^T  A^{hx}(t) z$ and
\begin{align*}
    &\la f(t) f(t') \ra - \la f(t) \ra \la f(t') \ra \\
    &\quad = b(t,t') + u^T B^{h}(t,t') u + z^T B^{x}(t,t') z
\end{align*}
where the expectation is over $p(f\cond K_u^{-1} u, K_z^{-1} z)$ and
\begin{align*}
    b(t,t) &= a(t,t') - \tr K_u^{-1} A^{h}(t,t') - \tr K_z^{-1}A^{x}(t,t') \\
    &\phan{=} + \tr K_u^{-1}A^{hx}(t)K_z^{-1}A^{xh}(t'), \\
    B^{h}(t,t') &= A^{h}(t,t') - A^{hx}(t)K_z^{-1} A^{xh}(t'), \\
    B^{x}(t,t') &= A^{x}(t,t') - A^{xh}(t)K_u^{-1} A^{hx}(t').
\end{align*}
Finally, we denote $a(t)=a(t,t)$ and do so for $A^h$, $A^x$, $A^{hx}$, $b$, $B^h$, and $B^x$ as well.

% \subsection{Causal Interdomain Transformation}
% In the case of the causal interdomain transformation, $A^{hx}$ and $A^x$ reduce to
% \begin{align*}
%    A^{hx}(t) &=\int^{t\land t_z^T} k_h(t_u,t-\tau) k_{xs}(\tau, t_z^T) \id{ \tau}, \\
%    A^{x}(t,t')&=\int^{t \land t_z}\!\!\!\!\int^{t'\land t_z^T} k_{sx}(t_z, \tau') k_h(t- \tau,t'-\tau') \\
%     &\phan{=\int^{t \land t_z}\!\!\!\!\int^{t'\land t_z^T}}\, k_{xs}(\tau, t_z^T) \id{\tau'}\id{\tau}.
% \end{align*}
